\section{Methods}\label{sec:methods}

\subsection{Problem Graph and Route Pool}\label{sec:graph}

We model the operational area as an undirected graph
\(G=(V,E)\) where \(|V| = n\) and node \(0\) is the depot.
An edge is created for every pair of customer nodes appearing in the
Delaunay triangulation of their planar coordinates; we then connect the
depot to \emph{every} customer, forming a sparse ``star–plus–local''
topology that matches typical drone-return logistics.
Thus connectivity is guranteed for a drone to return from any node to the depot,
yet we preserve realistic sparse topology. 
Each edge stores distance~\(d_{uv}\), travel
time~\(t_{uv}=d_{uv}/v_{d}\) (constant cruise speed~\(v_{d}\)),
battery usage~\(b_{uv}=t_{uv}\) and a boolean
\(c_{uv}=1\) if it belongs to a regulatory no-fly corridor.\footnote{%
Travel times and battery cost are identical in our instance; separating
them is straightforward.}

\vspace{2pt}\noindent
\textbf{Greedy soft-max route generator.}
Algorithm~1 (Appendix~A) repeatedly grows depot-to-depot paths using a
soft-max on incremental travel cost; feasibility is checked against
battery~\(B_{\max}\) and payload~\(Q_{\max}\).
After 2\,000 random seeds we retain the first
\(N=2\,048\) routes \(r\) that satisfy all constraints and pre-compute
\(\text{time}(r), \text{battery}(r)\) together with
node/edge bit-masks \(\boldsymbol{m}^V_r,\boldsymbol{m}^{E}_r\).
\emph{Feasibility is henceforth guaranteed};
the optimisation problem concerns \emph{only} combinatorial subset
selection.

% ------------------------------------------------------------------
\begin{algorithm}[H]
    \caption{GreedySoftmaxRouteGen}
    \label{alg:softmax}
    \begin{algorithmic}[1]
    \Require graph $G$, temperature $S$, battery $B_{\max}$, payload $Q_{\max}$
    \State $R \gets [\;]$
    \For{$run = 1,2,\dots$}
       \State $path \gets [0]$, $\textit{unvisited}\gets V\setminus\{0\}$
       \While{feasible node exists}
           \State compute $\Delta t$ and $\Delta b$ for each feasible next node
           \State draw next node with $\text{softmax}(-\Delta t/S)$
           \State update $path$, battery, payload
       \EndWhile
       \If{\Call{Feasible}{$path$}} \Comment{meets $B_{\max},Q_{\max}$}
            \State add new Route$(path)$ to $R$
       \EndIf
       \If{$|R|=2\,048$} \textbf{break}
    \EndFor
    \State \Return $R$
    \end{algorithmic}
\end{algorithm}
    

\subsection{Subset-Route Decision Space}

Selecting exactly \(k=10\) route–wave pairs
\(\mathcal{S}=\{(r_i,w_i)\}_{i=1}^{k}\)
with \(w_i\in\{0,\dots,M-1\}, M=4\) yields
\(\binom{N}{k}M^{k}\!\approx\! 2.6\times10^{26}\) possibilities.
We encode each pair by its \emph{index}:

% ------------------------------------------------------------------
% Eq. 2  (place in §2.2 Decision Encoding)
\begin{equation}
    \label{eq:logindex}
    r \;=\;
    \Bigl(\sum_{j=0}^{n_r-1}2^{j} z_j\Bigr)\bmod N,
    \qquad
    w \;=\;
    \Bigl(\sum_{j=0}^{n_w-1}2^{j} z_{n_r+j}\Bigr)\bmod M.
    \end{equation}
    

\noindent with \(n_r=\lceil\log_2 N\rceil=11\) and
\(n_w=\lceil\log_2 M\rceil=2\).
Classically this is a 130-bit chromosome; in the quantum circuit only
\(n_r+n_w=13\) \emph{qubits} are needed.

\vspace{2pt}\noindent
\textbf{Duplicate repair.}  
Within each wave we remove overlapping routes via the
first-occurrence rule:

% ------------------------------------------------------------------
\begin{algorithm}[H]
    \caption{GreedyDropDuplicates}
    \label{alg:dropdup}
    \begin{algorithmic}[1]
    \Procedure{DropDup}{$\mathcal{R}$}
      \State $covered \gets \mathbf{0}$ \Comment{bit‑mask of served nodes}
      \State $R' \gets [\;]$
      \ForAll{$r \in$ \Call{SortAsc}{$\mathcal{R}$, travel\_time}}
           \If{$\lnot(covered \;\&\; m^V_r)$}
               \State append $r$ to $R'$
               \State $covered \gets covered \;|\; m^V_r$
           \EndIf
      \EndFor
      \State \Return $R'$
    \EndProcedure
\end{algorithmic}
\end{algorithm}
    

\noindent The path of a retained route is \emph{not} modified; empirical
tests showed this suffices to achieve 100 \% node coverage once
penalties are set (Section~\ref{sec:cost}).

\subsection{Ascending-Time Wave Scheduler}

For each wave we assign routes to a drone min-heap in ascending
travel-time order (Algorithm~2).  
The classical longest-processing-time (LPT) variant is obtained by
reversing line 2; we leave this to future work.

% ------------------------------------------------------------------
\begin{algorithm}[H]
    \caption{AscendingTimeScheduler}
    \label{alg:scheduler}
    \begin{algorithmic}[1]
    \Procedure{Schedule}{$W$ waves, $D$ drones}
      \State $T_{\max}\gets0$, $A\gets[\;]$
      \ForAll{$w\;\textbf{in}$ \Call{SortedKeys}{$W$}}
          \State heapify($D$) \Comment{key = availability\_time}
          \ForAll{$r \in$ \Call{SortAsc}{$W[w]$, travel\_time}}
               \State $d \gets$ heappop($D$)
               \State \Call{Assign}{$d,r$} \Comment{updates $d.\text{end}$}
               \State $T_{\max}\gets\max(T_{\max}, d.\text{end})$
               \State append record to $A$; heappush($D,d$)
          \EndFor
      \EndFor
      \State \Return $(A, T_{\max})$
    \EndProcedure
\end{algorithmic}
\end{algorithm}
    

\subsection{Cost Oracle}\label{sec:cost}

Let \(\mathbb 1_{v\text{ served}}\) and
\(\mathbb 1_{uv\text{ traversed}}\) be binary indicators extracted from
bit-masks.
Our scalar objective mirrors the code:

% ------------------------------------------------------------------
% Eq. 1  (place in §2.4 Cost Oracle)
\begin{equation}
    \label{eq:cost}
    \boxed{
    C(\mathcal{S}) \;=\;
    10\,T_{\max}
    \;+\;
    \lambda_E \sum_{(u,v)\in E_c}
           \bigl(1-\mathbb 1_{uv}\bigr)^{\!2}
    \;+\;
    \lambda_V \sum_{v>0}
           \bigl(1-\mathbb 1_{v}\bigr)^{\!2}
    \;+\;
    \lambda_0 \sum_{v>0}
           \mathbb 1_{v}^{\mathrm{miss}}
    \;+\;
    \lambda_W \,
            \bigl(\text{overbook}\bigr)^{2}}
    \end{equation}
    

\noindent with default weights
\(\lambda_T\!=\!10,\;
 \lambda_V\!=\!10,\;
 \lambda_0\!=\!50\lambda_V,\;
 \lambda_E\!=\!10,\;
 \lambda_W\!=\!20\).
The high \(\lambda_0\) virtually prohibits missed deliveries; all best
solutions we report achieved 100 \% coverage.

\subsection{Variational Quantum Circuit}

We use the hardware-efficient
\(\mathrm{SU}(2)\) ansatz (Qiskit) on the 13-qubit register,

\[
U^{(L)}_{\boldsymbol{\theta}}
=\bigl[\;{\rm RY}(\theta)\;{\rm CZ}\bigr]^{L},
\quad L\in\{1,2\},
\]

\noindent sampled with \(N_{\text{shots}}=4\,096\) shots—within daily
budgets of current 127-qubit IBM back-ends.  Circuit depth remains
below 180 two-qubit gates for \(L=2\).

\subsection{Noise-Aware CVaR Objective}

The CVaR metric averages the best (lowest-cost) α-fraction of Monte-Carlo samples, 
following the noise-mitigation strategy of the CVaR-VQE proposal \cite{barkoutsos2020cvar}.

From shot counts we draw \(N_{\mathrm{mc}}=2\,000\) $k$-tuples and obtain
cost vector \(\boldsymbol{c}\).  The CVaR tail metric is

% ------------------------------------------------------------------
% Eq. 3  (place in §2.4 CVaR Objective)
\begin{equation}
    \label{eq:cvar}
    J_{\alpha}(\boldsymbol{c}) \;=\;
    \frac{1}{\alpha K}
    \sum_{i=1}^{\alpha K} c_{(i)},
    \quad
    0<\alpha\le 1,
\end{equation}
    

\noindent where \(c_{(i)}\) are costs sorted ascending.
Its estimator variance combines independent shot and Monte-Carlo terms:

% ------------------------------------------------------------------
% Eq. 4  (place just after Eq. 3)
\begin{equation}
    \label{eq:variance}
    \mathrm{Var}\!\bigl[J_{\alpha}\bigr]
    =\;\underbrace{\frac{\sigma_C^{2}}{N_{\text{mc}}}}_{\text{MC sampling}}
    \;+\;
    \underbrace{\sum_{i}
            \frac{p_i(1-p_i)}{N_{\text{shots}}}\,
            \bigl(C_i^{\mathrm{tail}}\bigr)^{2}}_{\text{shot noise}},
\end{equation}
    
\noindent
\textbf{Interpretation of the shot–noise term.}
Let $p_i=\Pr\!\bigl(\text{bitstring}=i\bigr)$ denote the probability of
measuring the $13$-qubit string ${i}$, which encodes the concatenation
$\ket{\texttt{route}_i}\ket{\texttt{wave}_i}$.  
Define
$C_i^{\mathrm{tail}} = C_i$ if the corresponding cost sample lies
within the \emph{best} $\alpha$-fraction used by CVaR, and
$C_i^{\mathrm{tail}} = 0$ otherwise.
Because $p_i$ already captures the joint chance of observing both a
route index \emph{and} its wave index, Equation~\eqref{eq:variance}
automatically accounts for the additional
``compression'' of two logical variables into a single bitstring
distribution; no extra factor is required.
Consequently CVaR implicitly \emph{down-weights} high-variance, low-%
probability strings—reducing the effective shot noise that drives the
optimiser.


\subsection{Bayesian Hyper-parameter Search}\label{sec:bo}

Parameter vectors \(\boldsymbol{\theta}\) are optimised with
Optuna TPE\cite{akiba2019optuna}.
We test two samplers:
(i) \emph{Default} (prior\_weight=1.0, $24$ EI samples) and
(ii) \emph{Aggressive} (prior\_weight=0.5, $64$ EI samples).
A MedianPruner terminates trials whose running best is above the median
of completed trials after one-third of the budget.
Optimiser budget is \(N_{\text{eval}}=100\) oracle calls—identical to
the GA baseline.

\paragraph{Tree–Parzen Estimator (TPE).}
TPE replaces Gaussian-process surrogate models with two
non-parametric density estimates:
$q_{\text{good}}(\boldsymbol{\theta})$ for parameters whose
objective lies in the top~$\gamma$ quantile,
and $q_{\text{bad}}(\boldsymbol{\theta})$ for the rest.
New trials are drawn to maximise the ratio
$q_{\text{good}}/q_{\text{bad}}$, an approximation of expected
improvement that \emph{does not assume homoscedastic noise}.
This is a natural match for our CVaR objective:
heteroscedastic shot noise is already damped by the CVaR tail
(Eq.\,\ref{eq:variance}), and TPE further biases sampling toward
regions where the \emph{distribution} of costs—not just the mean—is
promising.
We report results for both the Optuna defaults
(prior\_weight = 1.0, 24 EI candidates) and an ``aggressive’’ variant
(prior\_weight = 0.5, 64 candidates) that emphasises exploitation.


\subsection{Genetic Algorithm Baseline}

PyGAD with chromosome length 130 bits uses:
roulette-wheel selection, two-point crossover,
adaptive mutation \((p_{\text{high}},p_{\text{low}})=(0.06,0.012)\),
population size $5\times$ chromosome length
and enough generations to match the VQE’s
\(10^{5}\) oracle-call budget.
This configuration emerged as the best among several quick sweeps; a
detailed GA hyper-study is left to future work.

\subsection{Default Hyper-parameters}

\begin{table}[H]
\centering
\begin{tabular}{lcc}
\toprule
Parameter & Symbol & Value \\
\midrule
Battery capacity & $B_{\max}$ & 80 \\
Payload capacity & $Q_{\max}$ & 50 \\
Selected routes & $k$ & 10 \\
Circuit shots & $N_{\text{shots}}$ & 4\,096 \\
MC samples & $N_{\text{mc}}$ & 2\,000 \\
CVaR tail & $\alpha$ & 0.05 \\
Ansatz layers & $L$ & 1–2 \\
Oracle budget / trial &  & $10^{5}$ \\
\bottomrule
\end{tabular}
\caption{Default hyper-parameters shared across all experiments.}
\end{table}
